Testing Framework and Reference

### 2025-05-12 nbs/development/testing.qmd
docs(development): update testing docs
- note integration of bdd docs

Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Testing Framework
format:
  html:
    toc: true
---

## Overview

PyroVelocity uses a comprehensive testing framework to ensure code quality and correctness. This document describes the testing approach, tools, and guidelines for writing tests.

@@

The testing philosophy of PyroVelocity emphasizes:

- Creating simple instances of real objects over using mocking tools
- Writing tests alongside code development
- Ensuring tests provide reasonable constraints on the code
- Systematic validation of model implementations

## Test Organization

Tests are organized in the following structure:

- `src/pyrovelocity/tests/`: Root directory for all tests
  - `models/`: Tests for model components
  - `plots/`: Tests for plotting functions
  - `utils/`: Tests for utility functions
  - `data/`: Test data fixtures

@@

- Direct comparison of model outputs
- Validation using synthetic data from test fixtures
- Validation using downsampled real datasets
- Systematic addressing of implementation differences

## Writing Tests

When writing tests for PyroVelocity:

1. Use pytest fixtures for test data and common setup
2. Create simple instances of real objects rather than mocks
3. Test both normal operation and edge cases
4. Use xdoctests with proper type annotations
5. Ensure tests are self-contained with synthetic data

To run tests:

```bash
source .venv/bin/activate
pytest src/pyrovelocity/tests/
```

To run specific test modules:

```bash
source .venv/bin/activate
pytest src/pyrovelocity/tests/models/test_model.py
```

## Profiling and Benchmarking

@@

Tests are automatically run as part of the CI/CD pipeline defined in `.github/workflows/cid.yaml`.

>>>>

---
title: Testing Framework and Reference
format:
  html:
    toc: true
    toc-depth: 3
---

# PyroVelocity Testing

This documentation describes the testing approach and infrastructure for the `pyrovelocity` library. Our testing strategy combines traditional pytest unit/integration testing with Behavior-Driven Development (BDD) using pytest-bdd to ensure both technical correctness and alignment with user requirements.

## Directory Structure

The tests are organized to mirror the source code structure while separating unit tests from BDD feature definitions and step implementations:

```
src/pyrovelocity/tests/
├── __init__.py
├── conftest.py                # Shared pytest fixtures and configuration
├── fixtures/                  # Test data generation utilities
├── models/                    # Tests for model components
│   ├── __init__.py
│   ├── modular/               # Tests for modular implementation
│   │   ├── __init__.py
│   │   ├── conftest.py        # Modular-specific fixtures
│   │   ├── components/        # Unit tests for component implementations
│   │   │   ├── test_dynamics.py
│   │   │   ├── test_guides.py
│   │   │   ├── test_likelihoods.py
│   │   │   ├── test_observations.py
│   │   │   └── test_priors.py
│   │   ├── integration/       # BDD step implementations
│   │   │   ├── __init__.py
│   │   │   ├── conftest.py    # BDD-specific fixtures
│   │   │   ├── test_dynamics_model_steps.py
│   │   │   ├── test_model_steps.py
│   │   │   └── test_prior_model_steps.py
│   │   ├── selection/         # Tests for model selection
│   │   ├── utils/             # Tests for utility modules
│   │   ├── test_anndata_integration.py
│   │   ├── test_comparison.py
│   │   ├── test_factory.py
│   │   ├── test_interfaces.py
│   │   ├── test_model.py
│   │   ├── test_registry.py
│   │   └── test_unified_config.py
│   ├── jax/                   # Tests for JAX implementation (not yet documented)
│   └── test_*.py              # Tests for legacy implementation
├── features/                  # BDD feature specifications
│   └── models/
│       └── modular/
│           ├── dynamics_model.feature
│           ├── guide_model.feature
│           ├── likelihood_model.feature
│           ├── model.feature
│           ├── observation_model.feature
│           └── prior_model.feature
├── plots/                     # Tests for plotting functions
├── utils/                     # Tests for utility functions
├── validation/                # Tests for validation framework
└── test_*.py                  # Package-level tests
```

@@

Our testing approach implements a dual strategy that addresses both implementation details and user-facing behavior:

1. **Unit and Integration Tests** (`tests/models/modular/*.py`):
   - Traditional pytest-based tests focusing on implementation details
   - Verify internal functions, edge cases, and error handling
   - Target high code coverage and technical correctness
   - Validate that individual components work as expected in isolation

2. **Behavior-Driven Tests**:
   - **Feature Definitions** (`tests/features/models/modular/*.feature`):
     - User-centric specifications written in Gherkin syntax
     - Focus on requirements and use cases from the user's perspective
     - Serve as living documentation of expected behavior
   - **Feature Implementations** (`tests/models/modular/integration/*.py`):
     - Connect Gherkin specifications to actual code
     - Test how components work together to deliver user-facing functionality
     - Validate that the system behaves according to specifications

This approach follows Acceptance Test-Driven Development (ATDD), where feature files are written before implementation to document requirements and guide development.

### Key Principles

PyroVelocity's testing philosophy emphasizes:

- **Real Objects Over Mocks**: Creating simple instances of real objects rather than using mocking tools
- **Test-Driven Development**: Writing tests alongside code development
- **Reasonable Constraints**: Ensuring tests provide reasonable constraints on the code
- **Systematic Validation**: Comprehensive validation of model implementations
- **Protocol-First Testing**: Testing against interfaces rather than concrete implementations

### All Tests

```bash
source .venv/bin/activate
pytest src/pyrovelocity/tests
```

### Only Modular Implementation Tests

```bash
source .venv/bin/activate
pytest src/pyrovelocity/tests/models/modular
```

### Only BDD Tests

```bash
source .venv/bin/activate
pytest src/pyrovelocity/tests/models/modular/integration
```

### Specific Component Tests

```bash
source .venv/bin/activate
pytest src/pyrovelocity/tests/models/modular/components/test_dynamics.py
```

### Running with Coverage

```bash
source .venv/bin/activate
pytest src/pyrovelocity/tests --cov=pyrovelocity --cov-report=term-missing
```

### Profiling and Benchmarking

## Test Development Workflow

Our recommended workflow follows these steps:

1. **Specify Behavior**: Create or update a feature file in `features/` for new functionality
2. **Implement Step Definitions**: Create/update step definitions in the subpackage's `integration/` directory
3. **Add Unit Tests**: Create detailed unit tests in the subpackage's test modules
4. **Implement Source Code**: Develop the actual functionality in the source code
5. **Update Documentation**: Update this document's reference tables when adding new tests

@@

- Direct comparison of model outputs
- Validation using synthetic data from test fixtures
- Validation using downsampled real datasets (e.g., `tests/data/preprocessed_pancreas_50_7.json`)
- Systematic addressing of implementation differences

The validation framework is implemented in the `src/pyrovelocity/tests/validation/` directory and includes tests for:

- Direct comparison of model outputs
- Metrics for quantifying differences
- Visualization of differences

## Feature Specifications Reference

The table below tracks all feature specifications defined in our BDD feature files for the modular implementation. This serves as a living document of the intended functionality for the library.

| Module | Feature | Status | Description |
|--------|---------|--------|-------------|
| dynamics_model | Standard dynamics model computes expected counts | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes expected unspliced and spliced counts |
| dynamics_model | Standard dynamics model computes steady state | ✅ Implemented | Verify that the StandardDynamicsModel correctly computes steady state values |
| dynamics_model | Legacy dynamics model matches legacy implementation | ✅ Implemented | Verify that the LegacyDynamicsModel matches the legacy implementation |
| dynamics_model | Dynamics model handles edge cases | ✅ Implemented | Verify that the dynamics model handles edge cases gracefully |
| dynamics_model | Dynamics model with library size correction | ✅ Implemented | Verify that the dynamics model correctly applies library size correction |
| prior_model | LogNormal prior model creates stochastic nodes | ✅ Implemented | Verify that the LogNormalPriorModel creates appropriate stochastic nodes |
| likelihood_model | Poisson likelihood model creates observation nodes | ✅ Implemented | Verify that the PoissonLikelihoodModel creates appropriate observation nodes |
| likelihood_model | Legacy likelihood model matches legacy implementation | ✅ Implemented | Verify that the LegacyLikelihoodModel matches the legacy implementation |
| observation_model | Standard observation model preprocesses data | ✅ Implemented | Verify that the StandardObservationModel correctly preprocesses data |
| guide_model | Auto guide factory creates guide | ✅ Implemented | Verify that the AutoGuideFactory creates an appropriate guide |
| guide_model | Legacy auto guide factory matches legacy implementation | ✅ Implemented | Verify that the LegacyAutoGuideFactory matches the legacy implementation |
| model | PyroVelocity model integrates components | ✅ Implemented | Verify that the PyroVelocityModel correctly integrates all components |
| model | PyroVelocity model trains on AnnData | ✅ Implemented | Verify that the PyroVelocityModel can train on AnnData objects |

### Legend
- ✅ Implemented: Feature is fully implemented and tested
- ⚠️ In Progress: Implementation or testing is underway
- 📝 Specified: Feature is specified but not yet implemented
- 🔄 Planned: Feature is planned but not fully specified

## Unit Test Coverage Reference

This table tracks the unit test coverage for each module in the modular implementation, helping identify areas that need additional testing.

| Module | Test File | Test Coverage | Test Cases | Notes |
|--------|-----------|---------------|------------|-------|
| components/dynamics.py | test_dynamics.py | High | 10+ | Tests for StandardDynamicsModel and LegacyDynamicsModel |
| components/guides.py | test_guides.py | High | 5+ | Tests for AutoGuideFactory and LegacyAutoGuideFactory |
| components/likelihoods.py | test_likelihoods.py | High | 5+ | Tests for PoissonLikelihoodModel and LegacyLikelihoodModel |
| components/observations.py | test_observations.py | High | 5+ | Tests for StandardObservationModel |
| components/priors.py | test_priors.py | High | 5+ | Tests for LogNormalPriorModel |
| model.py | test_model.py | High | 10+ | Tests for PyroVelocityModel |
| factory.py | test_factory.py | High | 10+ | Tests for factory functions |
| registry.py | test_registry.py | High | 5+ | Tests for component registries |
| interfaces.py | test_interfaces.py | Medium | 5+ | Tests for Protocol interfaces |
| data/anndata.py | test_anndata_integration.py | High | 5+ | Tests for AnnData integration |
| utils/*.py | utils/test_*.py | Medium | 10+ | Tests for utility functions |

## BDD Feature Files Explained

Our feature files use Gherkin syntax to describe expected behavior. Each feature file follows this structure:

```gherkin
Feature: [Feature Name]
  As a [type of user]
  I want to [perform some action]
  So that [achieve some benefit]

  Background:
    Given [common setup steps]

  Scenario: [Specific use case name]
    Given [preconditions]
    When [actions]
    Then [expected results]
```

### Example Feature File

```gherkin
Feature: Dynamics Model
  As a computational biologist
  I want to model RNA velocity dynamics
  So that I can understand transcriptional dynamics in single-cell data

  Background:
    Given I have a dynamics model component
    And I have input data with unspliced and spliced counts

  Scenario Outline: Standard dynamics model computes expected counts
    Given I have a StandardDynamicsModel
    When I run the forward method with alpha <alpha>, beta <beta>, and gamma <gamma>
    Then the model should compute expected unspliced and spliced counts
    And the expected counts should follow RNA velocity dynamics

    Examples:
      | alpha | beta | gamma |
      | 1.0   | 0.5  | 0.2   |
```

## Step Definitions Structure

Step definition files connect feature files to actual test code. They are organized by component in the `integration/` directory:

```python
# src/pyrovelocity/tests/models/modular/integration/test_dynamics_model_steps.py
from importlib.resources import files
import pytest
import torch
from pytest_bdd import scenarios, given, when, then, parsers

# Import the feature file using importlib.resources
scenarios(str(files("pyrovelocity.tests.features") / "models" / "modular" / "dynamics_model.feature"))

# Import the components
from pyrovelocity.models.modular.components import (
    LegacyDynamicsModel,
    StandardDynamicsModel,
)

@given("I have a dynamics model component")
def dynamics_model_component():
    """Create a generic dynamics model component."""
    return StandardDynamicsModel()

@when(parsers.parse("I run the forward method with alpha {alpha}, beta {beta}, and gamma {gamma}"), target_fixture="run_forward_method_with_parameters")
def run_forward_method_with_parameters_fixture(standard_dynamics_model, input_data, alpha, beta, gamma):
    """Run the forward method with the given parameters."""
    # Convert string parameters to float
    alpha_val = float(alpha)
    beta_val = float(beta)
    gamma_val = float(gamma)

    # Create context with input data and parameters
    context = {
        "u_obs": input_data["u_obs"],
        "s_obs": input_data["s_obs"],
        "alpha": torch.tensor([alpha_val] * input_data["n_genes"]),
        "beta": torch.tensor([beta_val] * input_data["n_genes"]),
        "gamma": torch.tensor([gamma_val] * input_data["n_genes"]),
    }

    # Run the forward method
    result_context = standard_dynamics_model.forward(context)

    # Store the result for later steps
    return result_context
```

## Contributing New Tests

When contributing new tests to the project:

1. **Check Existing Coverage**: Review the feature and unit test reference tables
2. **Update Feature Files**: If implementing a new feature, add or update the relevant feature file
3. **Follow the Workflow**: Implement step definitions and unit tests as outlined above
4. **Run Tests Locally**: Ensure all tests pass before submitting a PR
5. **Update Documentation**: Update the reference tables in this document

## Fixtures and Testing Utilities

PyroVelocity provides a comprehensive set of fixtures for testing, defined in various `conftest.py` files:

### Main Fixtures (`src/pyrovelocity/tests/conftest.py`)

| Fixture Name | Purpose | Scope |
|--------------|---------|-------|
| `adata_preprocessed_pancreas_50_7` | Provides preprocessed pancreas dataset | Function |
| `adata_trained_pancreas_50_7` | Provides trained pancreas dataset | Function |
| `adata_postprocessed_pancreas_50_7` | Provides postprocessed pancreas dataset | Function |
| `default_sample_data` | Provides synthetic sample data | Function |
| `temp_file_path` | Provides a temporary file path | Function |
| `temp_compressed_pickle_path` | Provides a temporary compressed pickle path | Function |
| `save_and_load_helper` | Helper for saving and loading data | Function |

### Modular Implementation Fixtures (`src/pyrovelocity/tests/models/modular/conftest.py`)

These fixtures are specific to the modular implementation and provide test instances of various components.

### BDD-Specific Fixtures (`src/pyrovelocity/tests/models/modular/integration/conftest.py`)

| Fixture Name | Purpose | Scope |
|--------------|---------|-------|
| `bdd_simple_data` | Provides simple data for BDD testing | Function |
| `bdd_model_parameters` | Provides model parameters for BDD testing | Function |
| `bdd_standard_dynamics_model` | Provides a StandardDynamicsModel | Function |
| `bdd_legacy_dynamics_model` | Provides a LegacyDynamicsModel | Function |
| `bdd_lognormal_prior_model` | Provides a LogNormalPriorModel | Function |
| `bdd_poisson_likelihood_model` | Provides a PoissonLikelihoodModel | Function |
| `bdd_legacy_likelihood_model` | Provides a LegacyLikelihoodModel | Function |
| `bdd_standard_observation_model` | Provides a StandardObservationModel | Function |
| `bdd_auto_guide_factory` | Provides an AutoGuideFactory | Function |
| `bdd_legacy_auto_guide_factory` | Provides a LegacyAutoGuideFactory | Function |
| `bdd_pyro_velocity_model` | Provides a PyroVelocityModel | Function |
| `bdd_anndata` | Provides a simple AnnData object | Function |
| `clear_pyro_param_store` | Clears Pyro's parameter store | Function (autouse) |

## Advanced Testing Topics

### Parameterized Testing

Both unit tests and BDD scenarios can be parameterized:

```python
# Unit test parameterization
@pytest.mark.parametrize("alpha,beta,gamma,expected_u_ss,expected_s_ss", [
    (1.0, 0.5, 0.2, 2.0, 5.0),
    (2.0, 1.0, 0.5, 2.0, 4.0),
])
def test_steady_state(alpha, beta, gamma, expected_u_ss, expected_s_ss):
    model = StandardDynamicsModel()
    u_ss, s_ss = model.steady_state(
        torch.tensor([alpha]),
        torch.tensor([beta]),
        torch.tensor([gamma]),
    )
    assert torch.allclose(u_ss, torch.tensor([expected_u_ss]))
    assert torch.allclose(s_ss, torch.tensor([expected_s_ss]))
```

```gherkin
# BDD parameterization with Scenario Outline
Scenario Outline: Standard dynamics model computes steady state
  Given I have a StandardDynamicsModel
  When I compute the steady state with alpha <alpha>, beta <beta>, and gamma <gamma>
  Then the steady state unspliced should equal alpha/beta
  And the steady state spliced should equal alpha/gamma

  Examples:
    | alpha | beta | gamma |
    | 1.0   | 0.5  | 0.2   |
    | 2.0   | 1.0  | 0.5   |
```

### Testing with AnnData

PyroVelocity provides fixtures for testing with AnnData objects:

```python
def test_anndata_integration(bdd_anndata, bdd_pyro_velocity_model):
    """Test integration with AnnData."""
    # Prepare AnnData
    adata = bdd_anndata.copy()

    # Train model
    bdd_pyro_velocity_model.train(
        adata,
        max_epochs=2,
        batch_size=5,
        use_gpu=False,
    )

    # Check that results were stored in AnnData
    assert "dynamics" in adata.uns
    assert "model_type" in adata.uns["dynamics"]
```

### Testing with Pyro

PyroVelocity provides utilities for testing with Pyro:

```python
def test_pyro_model(clear_pyro_param_store):
    """Test a Pyro model."""
    # Define a simple model
    def model():
        x = pyro.sample("x", pyro.distributions.Normal(0, 1))
        return x

    # Define a guide
    def guide():
        loc = pyro.param("loc", torch.tensor(0.0))
        scale = pyro.param("scale", torch.tensor(1.0), constraint=pyro.constraints.positive)
        pyro.sample("x", pyro.distributions.Normal(loc, scale))

    # Run inference
    pyro.clear_param_store()
    svi = pyro.infer.SVI(
        model=model,
        guide=guide,
        optim=pyro.optim.Adam({"lr": 0.01}),
        loss=pyro.infer.Trace_ELBO(),
    )

    # Train for a few steps
    for _ in range(10):
        loss = svi.step()

    # Check that parameters were updated
    assert pyro.param("loc").item() != 0.0
```

@@

Tests are automatically run as part of the CI/CD pipeline defined in `.github/workflows/cid.yaml`. The CI pipeline includes:

1. **Unit Tests**: Run all unit tests with coverage reporting
2. **BDD Tests**: Run all BDD tests to verify behavior
3. **Coverage Enforcement**: Ensure test coverage meets minimum thresholds

## Best Practices for PyroVelocity Testing

1. **Use Protocol Interfaces**: Test against Protocol interfaces rather than concrete implementations
2. **Create Real Objects**: Prefer creating simple instances of real objects over using mocks
3. **Test Edge Cases**: Include tests for edge cases and error handling
4. **Use Fixtures**: Leverage pytest fixtures for common setup and teardown
5. **Clear Pyro State**: Always clear Pyro's parameter store before and after tests
6. **Synthetic Data**: Use synthetic data for tests to avoid external dependencies
7. **Parameterize Tests**: Use parameterization to test multiple cases with the same logic
8. **Document Tests**: Include docstrings and comments to explain test purpose and logic
9. **Follow BDD Patterns**: Use Given-When-Then pattern for BDD tests
10. **Validate Against Legacy**: Validate new implementations against legacy code
